---
title: "Times Series - pilot on anomaly detection"
author: "Cary"
date: "9/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load the IR page-click data 
```{r}
montana <- read.csv('montana_state_university_RAMP_pc_daily_clicks.csv')
head(montana)
```

# Review data type 
```{r}
str(montana)
```

# Convert date to standard date class in R format

```{r}
montana$date <- as.Date(montana$date)
# as.Date(montana$date, "%m/%d/%Y") if the input file have a different format
```
# Drop repository_id
```{r}
montana <- subset(montana, select = -repository_id)
```

```{r}
head(montana)
```
Date has to be sorted first. 

```{r}
library(lubridate)
library(dplyr)
```

```{r}
montana <- montana %>% arrange(montana$date)
```

```{r}
str(montana)
```

# Method 1: use base R ts() function and plot.ts() 

```{r}
montana_series <- ts(montana) 
```

```{r}
class(montana_series)
```


```{r}
plot.ts(montana_series)
```
The ts() function is very limited in terms of handling daily values in our usage data. 

# Method 2: using the logic in R Statistics Cookbook, Chapter 7 by Juretig, 2019, on anomaly detection
```{r}
library(tscount)
library(dplyr)
library(anomalize)
library(tibbletime)
```
## Calculate days difference from the date column. 
```{r}
difference = montana %>% 
  mutate(diff_days = as.numeric(date-lag(date))) #append a new column diff_days by subtracting the date difference
```

```{r}
difference[is.na(difference)]  = 0  #Convert all na values in dataframe to zero 
```

```{r}
head(difference)
```

## Anomaly Detection 
### Transform the dataset into tibbletime object and assigned date as index 
```{r}
montana_anomalies = as_tbl_time(difference,index = date) 
str(montana_anomalies)
```

```{r}
head(montana_anomalies)
```
### Decompose the time series into trend, seasonality, and remainder components 
We assumed the IR usage data is seasonal. Decomposition will estimate the trend component, seasonal component, and irregular component 

```{r}
montana_anomalies %>% 
  time_decompose(clicks, merge = TRUE) #time_decompose from anomalize library 
# It gives you an output of how many months in frequency and trend calculation
# Four columns are displayed for time tibble: observed, season, trend, and remainder 
```
### Detect the anomalies using the remainder
```{r}
montana_anomalies %>% 
  time_decompose(clicks, merge = TRUE)%>%
  anomalize(remainder, method = "iqr")
```

### Putting the code sequence together and assigned to dataframe 

```{r}
results = 
  montana_anomalies %>%
  time_decompose(clicks, method = "stl", merge = TRUE) %>%
  anomalize(remainder, method = "iqr")
```
```{r}
head(results)
```

### Finally, join everything back together 

```{r}
montana_result <- results%>% time_recompose()
```

### Plot the result 
```{r, fig.width = 10, fig.asp = 1}
montana_result %>% plot_anomaly_decomposition(ncol=2,alpha_dots = 0.3) 
```

